A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-11-09 analysis/paper/paper.Rmd
Attempting to use Dockerfile

The small discrepancies between the achieved results mainly follow from the fact that a slightly updated version of the local displacement curve is applied here [cf. @sørensen2014; @sorensen2023]. Jaksland's dates are given a flat 200- and 50-year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while his argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they do not overlap. This information is not integrated in the approach outlined here, but it could justify further reducing the uncertainty ranges.

Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does result in very precise dates. Above it was suggested that additional temporal data could be combined with the method to improve its precision. Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use. One potential way to do this could be through the analysis of phosphate concentrations in soils, which has the potential to offer insights on the likely position of the shoreline when a site was in use [@ilves2011]. This has been done in the Baltic sea region [e.g. @broadbent1979; @sundström2006; @ilves2011], but has yet to provide reliable results in Norway [e.g. @melvold2014a; @viken2018b]. The identification of other physical traces of shore formation processes and the deposition of beach sediments in relation to archaeological material also holds similar potential [e.g. @bondevik2019]. Finally, another approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline. If such methods are developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use. Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf. Figure \@ref(fig:redate)).

@@

Other factors related to the topographic location of the sites could also be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled. This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range and potential impact of storm surges could also have implications for the location of a site relative to the shoreline, depending on the topography [@helskog1978; @bondevik2019]. The potential of exploring such dimensions was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline. If patterns related to such locational patterns can be discerned and unpacked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available. A mention should also be made here of the fact that catastrophic events such as tsunamis might also be of relevance [e.g. @blankholm2020; @nielsen2020; @nyland2021]. Evidence for the impact of tsunamis in the Stone Age has not been identified in south-eastern Norway as of yet [see @romundset2015, 398], and might therefore not be of direct relevance to the coastal settlement in the region. However, the outburst flood resulting from the catastrophic drainage of the glacial lake Nedre Glomsjø around 8500--8000 BCE [@høgaas2019], located in Mid-Norway some 230 km north of present-day Oslo, could have had consequences for how the coast was utilised [@solheim2020b, 9].

>>>>

The small discrepancies between the achieved results mainly follow from the fact that a slightly updated version of the local displacement curve is applied here [cf. @sørensen2014; @sorensen2023]. Jaksland's dates are given a flat 200- and 50-year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while his argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they do not overlap. This information is not integrated in the approach outlined here, but it could justify further reducing the uncertainty ranges. Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates.

Above it was suggested that additional temporal data could be combined with the method to improve its precision. Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use. One potential way to do this could be through the analysis of phosphate concentrations in soils, which has the potential to offer insights on the likely position of the shoreline when a site was in use [@ilves2011]. This has been done in the Baltic sea region [e.g. @broadbent1979; @sundström2006; @ilves2011], but has yet to provide reliable results in Norway [e.g. @melvold2014a; @viken2018b]. The identification of other physical traces of shore formation processes and the deposition of beach sediments in relation to archaeological material also holds similar potential [e.g. @bondevik2019]. Finally, another approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline. If such methods are developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use. Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDRs are already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf. Figure \@ref(fig:redate)).

@@

Other factors related to the topographic location of the sites could also be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled. This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range and potential impact of storm surges could also have implications for the location of a site relative to the shoreline, depending on the topography [@helskog1978; @bondevik2019]. The potential of exploring such dimensions was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline. If patterns related to such locational patterns can be discerned and unpacked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available. A mention should also be made here of the fact that catastrophic events such as tsunamis might also be of relevance [e.g. @blankholm2020; @nielsen2020; @nyland2021]. Evidence for the impact of tsunamis in the Stone Age has not been identified in south-eastern Norway as of yet [see @romundset2015, 398], and might therefore not be of direct relevance to the coastal settlement in the region. However, the outburst flood resulting from the catastrophic drainage of the glacial lake Nedre Glomsjø around 8500--8000 BCE [@høgaas2019], located in Mid-Norway some 230km north of present-day Oslo, could have had consequences for how the coast was utilised [@solheim2020b, 9].
